# Data Visualization Principles

1. Pie charts ar appropriate when

  A. When we want to display percentages
  B. When `ggplot2` is not available
  C. When I am in a bakery
  D.  Never. Barplots and tables are always better.
   

2. What is the problem with the plot below:

```{r echo=FALSE}
data.frame(candidate=c("Clinton","Trump"), electoral_votes = c(232, 306)) %>%
  ggplot(aes(candidate, electoral_votes)) +
  geom_bar(stat = "identity", width=0.5, color =1, fill = c("Blue","Red")) +
 coord_cartesian(ylim=c(200,310)) +
  ylab("Electoral Votes") +
  xlab("") +
  ggtitle("Result of Presidential Election 2016")
```

  A. The values are wrong. The final vote was 306 to 232.
  B. The axis does not start at 0. Judging by the length, it appears Trump recieved 3 times as many votes when in fact it about 30% more. 
  C. The colors should be the same.
  D. Percentages should be shown as pie chart
  
  
3. Take a look at the following two plots. They show the same information: 1928 rates of Measels across the 50 states

```{r}
library(gridExtra)
data("us_contagious_diseases")
p1 <- us_contagious_diseases %>% 
  filter(year == 1928 & disease=="Measles" & count>0 & !is.na(population)) %>% 
  mutate(rate = count / population * 10000) %>%
  ggplot(aes(state, rate)) +
  geom_bar(stat="identity") +
  coord_flip() +
  xlab("")
p2 <- us_contagious_diseases %>% 
  filter(year == 1928 & disease=="Measles" & count>0 & !is.na(population)) %>% 
  mutate(rate = count / population * 10000) %>%
  mutate(state = reorder(state, rate)) %>%
  ggplot(aes(state, rate)) +
  geom_bar(stat="identity") +
  coord_flip() +
  xlab("")
grid.arrange(p1, p2, ncol = 2)
```

Which one plot do is easier to read if you are interest and best and worst states and why?

    A. They provide the same information so they are just as good.
    B. The plot on the right is better because it orders the states alphabetically.
    C. The plot on the right is better because alphebatical order has nothing to do with the disease and by order by the acutal rate we quickly see the states with most and least rates.
    D. Both plots should be a piechart.
    
4. To make the plot on the left we have to reorder the levels of the states variable

```{r}
dat <- us_contagious_diseases %>%  filter(year == 1928 & disease=="Measles" & count>0 & !is.na(population))
```

Note what happens when we make a barplot:

```{r}
dat %>% ggplot(aes(state, rate)) +
  geom_bar(stat="identity") +
  coord_flip() 
```

Define these objets

```{r}
state <- dat$state
rate <- dat$count/dat$population*10000
```

Redefine the `state` object so that the levels are re-ordered. Print the new object `state` and it's levels so you can see that the vector is not re-ordered by the levels.

5. Now with one line of code, define the `dat` table as done above, but change the use mutate to create a rate variable and reorder state variable so that the levels are reorders by this variable. The make a baroplot using the code above, but for this new dat.

```{r}
dat <- us_contagious_diseases %>%  filter(year == 1928 & disease=="Measles" & count>0 & !is.na(population)) %>%
  mutate(res = count/population*10000) %>%
  mutate(state = reorder(state, res))

dat %>% ggplot(aes(state, rate)) +
  geom_bar(stat="identity") +
  coord_flip() 
```

5 - Say we are interested in comparing the gun homicide rates across regions of the US. We see this plot


```{r}
library(dslabs)
data("murders")
murders %>% mutate(rate = total/population*100000) %>%
  group_by(region) %>%
  summarize(avg = mean(rate)) %>%
  mutate(region = factor(region)) %>%
  ggplot(aes(region, avg)) +
  geom_bar(stat="identity") +
  ylab("Murder Rate Average")
```

and decide to move to a state in the West region. What is the main problem with this interpretaion.

  A. The categories are ordered alphabetically.
  B. The graph does not show standard errors.
  C. It does not show all the data. We do not see the variability withing region and it's possible that the safest states are not in the West.
  D. The Northeast has the lowest average.

6. Make a boxplot of the murder rates

```{r}
data("murders")
murders %>% mutate(rate = total/population*100000) 
```

by region, showing all the points and ordering the regions by their median rate.

  
7. This plots below show the for three continuos variables. 

```{r}
library(scatterplot3d)
library(RColorBrewer)
rafalib::mypar(1,1)
set.seed(1)
n <- 25
group <- rep(1,n)
group[1:(round(n/2))] <- 2
x <- rnorm(n, group, .33)
y <- rnorm(n, group, .33)
z <- rnorm(n)

scatterplot3d(x,y,z, color = group, pch=16)
abline(v=4, col=3)
```

The line $x=2$ appears to separate the points. But it is actually not the case, which we can see by plotting the data in a couple of two dimensinal points.
```{r}
mypar(1,2)
plot(x,y, col=group, pch =16)
##abline(3,-1,col=4)
abline(v=2, col=3)
plot(x,z,col=group, pch=16)
abline(v=2, col=3)
```

Why is this happening?

  A. Humans are not good at reading pseado 3D plots
  B. There must be  an error in the code
  C. The colors confuse us.
  D. Scatter-plots should not be used to compare two variables when we have access to 3.
