## Indexing


R provides a powerful and convinient way of indexing vectors. We can, for example, subset a vector based on properties of another vector. We continue our us murders example to demonstrate

### Subsetting with logicals

We have now calculated the murder rate using

```{r}
murder_rate <- murders$total / murders$population * 100000 
```

Say you are moving from Italy where, according to the ABC news report, the murder rate is only 0.71 per 100,000. You would prefer to move to a state with a similar rate. Another powerful feature of R is that we can we can use logicals to index vectors. 
Note that if we compare a vector to a single number, it actually performs the test for each entry. Here is an example related to the question above

```{r}
ind <- murder_rate < 0.71
```

Or if we want to know if its less or equal we can use

```{r}
ind <- murder_rate <= 0.71
ind
```

Note that we get back a  logical vector with `TRUE` for each entry smaller or equal than 0.71. To see which states these are, we can leverage the fact that vectors can be indexed with logicals.

```{r}
murders$state[ind]
```

Note that to count how many are TRUE, the function `sum` returns the sum of the entries of a vector and logical vectors get _coerced_ to numeric with `TRUE` coded as 1 and `FALSE` as 0. Thus we can count the states using

```{r}
sum(ind)
```


### Logical Operators

Suppose we like the mountains and we want to move to a safe state in the West region of the country. We want the murder rate to be at most 1. So we want two different things to be true. Here we can use the logical operator _and_ which in R is `&`. This operation results in a true, only when both logicals are true. To see this consider this example

```{r}
TRUE & TRUE
TRUE & FALSE
FALSE & FALSE
```

We can form two logicals:

```{r}
west <- murders$region=="West"
safe <- murders$rate<=1
```

and we can use the `&`  to get a vector of logicals that tells us which states satisfy our condition

```{r}
ind <- safe & west
murders$state[ind]
```

### `which`

Suppose we want to look up California's murder rate. For this type of operation, it is convenient to convert vectors of logicals into indexes instead of keeping long vectors of logicals. The function `which` tells us which entries of a logical vector are TRUE. So we can type:

```{r}
ind <- which(murders$state == "California")
ind ##this is the index that matches the California entry
murder_rate[ind]
```


### `match`

If instead of just one state we want to find out the murder rates for several, say New York, Florida, and Texas, we can use the function `match`. This function tells us which indexes of a second vector match each of the entries of a first vector:

```{r}
ind <- match(c("New York", "Florida", "Texas"), murders$state)
ind
```

Now we can look at the murder rates:

```{r}
murder_rate[ind]
```

### `%in%`

If rather than an index we want a logical that tells us whether or not each element of a first vector is in a second we can use the function `%in%`. So, say you are not sure if Boston, Dakota and Washington are states, you can find out like this

```{r}
c("Boston", "Dakota", "Washington") %in% murders$state
```


**Advanced**: Note that there is a connection between `match` and `%in% through `which`: the following two lines of code are equivalent:

```{r}
match(c("New York", "Florida", "Texas"), murders$state)
which(murders$state%in%c("New York", "Florida", "Texas"))
```

