## Models

1. We have been using urn models to motivate the use of probability models. Most data science applications are not related to data obtained from urns. More common are data that come from individuals. The reason probability plays a role here is because the data come from a random sample. The random sample is taken from a population. The urn serves as an analogy for the population. 

Let's revisit the heights dataset. Suppose we consider the males in our course the population.

```{r}
library(dplyr)
library(dslabs)
data(heights)
x <- heights %>% filter(sex == "Male") %>%
  .$height
```

Mathematically speaking, `x` is our population. Using the urn analogy, we have an urn with the values of  `x` in it. What are the population average and standard deviation of our population?


2.  Call the population average computed above $\mu$ and the standard deviation $\sigma$. Now take a sample of size 50, with replacement, and construct an estimate for $\mu$ and $\sigma$. Set the seed at 1 based on what has been described in this section.


3. What does the theory tell us about the sample average $\bar{X}$ and how it related to $\mu$? 

  A. It is practically identical to $\mu$.
  B. It is a random variable with expected value $\mu$ and standard error $\sigma/\sqrt{N}$
  C. It is a random variable with expected value $\mu$ and standard error $\sigma$.
  D. Contains no information
  
  
4. So how is this useful? We are going to use an over-simplified yet illustrative example. Suppose we want to know the average height of our male students but we only get to measure 50 of the 708. We will use $\bar{X}$ as our estimate. We know from the answer to 3 that the standard estimate of our error $\bar{X}-\mu$ is $\sigma/sqrt{N}$. If want to know what this is, but we don't know $\sigma$. Based on what is described in thie section, show your estimate of $\sigma$


5. Now that we have an estimate of $\sigma$, let's call our estimate $s$. Construct a 95% confidence interval for $\mu$.



6. Now run a Monte Carlo simulation in which you 
compute 10,000 confidence intervals as you have just done. What proportion of these intervals include $\mu$? Set the seed to 1.


7. In this section we talked about pollster bias. We used visualization to motivate the presence of such bias. Here we will give it a more rigorous treatment. We will consider two pollsters that conducted daily polls. We will look at national polls for the month before the election. 

```{r}
data(polls_us_election_2016)
polls <- polls_us_election_2016 %>% 
  filter(pollster %in% c("Rasmussen Reports/Pulse Opinion Research","The Times-Picayune/Lucid") &
           enddate >= "2016-10-15" &
           state == "U.S.") %>% 
  mutate(spread = rawpoll_clinton/100 - rawpoll_trump/100) 
```

We want to answer the quetion "Is there a poll bias?". Make a plot showing the spreads for each poll.



8. The data does seem to suggest there is a difference. But these
data are subject to variability. Maybe the differences we observe are due to chance. 

  The urn model theory tells says nothing about pollster effect. Under the urn model both pollsters have the same expected value: the election day difference, that we call $d$. 
  
  To answer this question, is there an urn model, we will model the observed data $Y_ij$ in the following way

$$
Y_{ij} = d + b_i + \varepsilon_{ij}
$$
with $i=1,2$ indexing the two pollsters, $b_i$ the bias for pollster $i$ and $\varepsinon_ij$ poll to poll chance variability. We assume the $\varepsilon$ are independent from each other, have expected value $0$ and standard deviation $\sigma_i$ regardless of $j$. 

Which of the following best represents our question?

  A. Is $\varepsilon_ij$ = 0
  B. How close are the $Y_ij$ to $d$
  C. Is $b_1 \neq b_2$?
  D. Are $b_1 = 0$ and $b_2 = 0$ ?
  
9. Note thet in the right side of this model only $\varepsilon_ij$ is a random variable. The other two are constants. What is the expected value of $Y_{1j}$?


10. Suppose we define $\bar{Y}_1$ as the average of poll results from the first poll, $Y_{11},\dots,Y_{1N_1}$ with $N_1$ the number of polls conducted by the first pollster:

```{r}
polls %>% 
  filter(pollster=="Rasmussen Reports/Pulse Opinion Research") %>% 
  summarize(N_1 = n())
```

What is the expected values $\bar{Y}_1$?


11. What is the standard error of $\bar_{Y}_1$ ?


12. What is the expected values $\bar{Y}_2$?


13. What is the standard error of $\bar_{Y}_2$ ?


14. Using what we learned in answering the questions above what is the expected value of $\bar{Y}_{2} - \bar{Y}_1$?


15. Using what we learned in answering the questions above what is the standard error of $\bar{Y}_{2} - \bar{Y}_1$?

